Automatic Parallelization and Transparent Fault Tolerance

### 2016-07-12 pubs/tfp16-draft/paper.tex
rename dir

\renewcommand{\floatpagefraction}{0.8}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (\emph{Project article})}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}%\thanks{Ringo, Anderson, and Marks are undergraduate students.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\hspace{0.2in} Our development has been greatly informed by the extant
literature and we indicate both where we follow, and deviate from,
these sources.
%
\\
{\color{OliveGreen} % \bf
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
\item We are reporting work in progress.  At the time of writing we have a
  serial implementation, a primitive proof-of-concept parallel
  implementation, and are currently developing a more realistic parallel
  runtime.  
%We expect to have preliminary results from our new parallel
%implementation by the time of the symposium.  
The fault-tolerance mechanisms remain future work.

\item Hopefully in the spirit of a student-friendly symposium, this draft is
  liberally annotated with indicators of student contributions, student
  preparation, the influence of their involvement on our approach, and other
  incidental and ancillary comments.  A final paper would of course be cleaned
  up and potentially considerably shortened.

\item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
  (all undergraduate) have contributed considerably to this effort, and to the
  writing of this paper, and will continue to do so.  If this paper is
  accepted we plan to have one of these students present it at TiFP'16.
\end{enumerate}
}

@@

A minor, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness and laziness.
%\footnote{The starting experience of our student interns to
%  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
%    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
%  substituting his article \emph{Higher-order Functions for
%    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
%  book.}

@@

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
library~\cite{ghc-as-library}.  Initially we start with our own STG.
%, with bridging the gap with GHC scheduled to commence quite soon.

\subsection{STG Language}

@@

\subsection{Degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term lazy and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

\section{Current Status (April 8, 2016), Future Work}

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the minuscule budget (and therefore the employment of student interns) we must
maintain sharp focus on the main goals: implementing and evaluating mechanisms
for automatic parallelism and fault tolerance.  At the time of writing we have
a fully functioning serial implementation including
\begin{itemize}
\item STG parser \emph{(by student)},
\item mini-Haskell parser, mini-Haskell to STG transformer \emph{(by student)},
\item type inferencer,
\item various analysis and transformation passes \emph{(some by students)},
\item code generator \emph{(some student contribution)},
\item runtime system including a garbage collector,
\item continuous integration and testing system,
\item extensive test suite \emph{(mostly by students)}.\footnote{Many are
    students' solutions to exercises in learning algorithms and data
    structures.}
\end{itemize}

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.

This summer (2016) a returning student will bridge the gap between GHC STG (or
Core) and our STG, augmenting the runtime as needed.  Another student will
develop an instrumentation infrastructure integrated with the testing
framework to record and present time and space run-time statistics such as
number of heap and stack allocations and maximum stack depth.

%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

% because # looks strangely large next to math terms
\newcommand{\hash}{{\scriptsize\#}}

\begin{figure}
\centering
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}
Variable         & $f,\ x$             &     &  & Initial lower-case letter \\

@@

\section{Student Preparation}

Student interns' backgrounds in computer programming has ranged from minimal
to some exposure to one or more mainstream imperative languages, Lisp, and
Haskell.  To bring them to a normalized level of proficiency their first
assignment has been to work through the entirety of Hutton's \emph{Programming
  in Haskell} textbook~\cite{Hutton-book}, substituting one of his articles
\emph{Higher-order Functions for Parsing}~\cite{Hutton-parsing:1992} or
\emph{Monadic Parser Combinators}~\cite{Hutton:1996} for the corresponding
material in the book.  This may be followed by writing a parser and ADT for a
progressively enriched lambda calculus, together with call-by-value,
call-by-name, and (in one case) call-by-need interpreters.  The concept of
compilation can be illustrated with an exercise in parsing infix arithmetic
expressions to an AST, and transforming that to a sequence of operations for a
simple stack machine.  From this point they have then been able to substantively
contribute to this project.

% breaks page limit
%  This learning phase can take the
%better part of a summer or one-semester internship, so we particularly
%seek return interns.

%
% ---- Bibliography ----
%

>>>>

\renewcommand{\floatpagefraction}{0.99}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}%\thanks{Ringo, Anderson, and Marks are undergraduate students.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\hspace{0.2in} Our development has been greatly informed by the extant
literature and we indicate both where we follow, and deviate from,
these sources.

@@

A minor, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of non-strictness.

@@

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
library~\cite{ghc-as-library}.  Initially we started with our own STG.

\subsection{STG Language}
\renewcommand{\floatpagefraction}{0.99}

@@

\subsection{Degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term lazy and simply refer to
strictness properties.\footnote{%
  We can readily simulate the absence of sharing (i.e, call by name instead of
  call by need), in the serial case at least, simply by not updating \emph{THUNK}s
after evaluation, resulting in, for example, exponential running time for
simple mergesort on lists.}
  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

\section{Current Status, Future Work}

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the minuscule budget (and therefore the employment of student interns) we must
maintain sharp focus on the main goals: implementing and evaluating mechanisms
for automatic parallelism and fault tolerance.  Currently we have
a fully functioning serial implementation including
\begin{itemize}
\item STG parser,
\item mini-Haskell parser, mini-Haskell to STG transformer,
\item type inferencer,
\item various analysis and transformation passes,
\item code generator,
\item runtime system including a garbage collector,
\item continuous integration and testing system,
\item extensive test suite, automated testing.
\end{itemize}

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
Bridging the gap between GHC STG and our STG, and the corresponding
augmentation of the runtime is showing early signs of success.
%
An instrumentation infrastructure, integrated with the testing framework to
record and present time and space run-time statistics such as number of heap
and stack allocations and maximum stack depth, is planned for the near future.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.
%
%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

% because # looks strangely large next to math terms

\newcommand{\hash}{{\scriptsize\#}}
\begin{figure}[h!tb]
\centering
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}
Variable         & $f,\ x$             &     &  & Initial lower-case letter \\

@@
